Beyond May: Complexity-stability relationships in high-dimensional dynamical systems

### 2023-03-15 papers/general-stability.tex
not much going on

\begin{abstract}
Fifty years ago, Robert May predicted that large, complex systems are unlikely to be stable. Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems: in some cases, strong interaction and large diversity are, indeed, destabilizing; in other cases, they are stabilizing. Which class a given system falls into depends on the relative convexity of the response to self- vs. cross-interactions. We illustrate the transition between "May" (complexity begets instability) and "anti-May" (complexity begets stability) behavior with a non-linear  generalization of the Lotka-Volterra model. 
\end{abstract}

@@

In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients: spatial or food-web structure, time delays, etc. Here, we show that May's argument is incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.

@@

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the contribution of a population $i$ to its own growth (its "production function"), while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. In the following we call $g(x_i)$ the "response function".

The classic generalized Lotka-Volterra competition model corresponds to $f$, $g$, $h$ all linear. However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$. Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks. For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level. It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.

>>>>

\begin{abstract}
Fifty years ago, Robert May predicted that large, complex systems are unlikely to be stable. Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems, depending on the relative convexity of the response to self- vs. cross-interactions. We illustrate the transition between "May" (complexity begets instability) and "anti-May" (complexity begets stability) behavior with a non-linear generalization of the Lotka-Volterra model. 
\end{abstract}

@@

In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc. 
Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.

@@

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.

The classic generalized Lotka-Volterra (GLV) competition model corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. 
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$. 
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks. 
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level. 
It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.
